Foreground Signatures and Mitigation in Neutral Hydrogen Power Spectra From The Reionization Epoch

### 2014-10-17 foreground_delay_spectrum_modeling.tex
From 89b3f2f: ============
(*) Revised section on delay spectrum math and changed notations.
(*) Added a few references in eor.bib

Considerable effort has been made towards understanding the $k$--space behavior of foreground signatures in the observed power spectrum \citep{thy13,pob13,mor12,tro12,liu11,dat10,bow09,liu09}. A general model of scattering of smooth spectrum foregrounds to higher $k$--modes by the wide-field response of the instrument, into the so-called `wedge', has emerged that provides a high level explanation for the observed foreground power spectrum. The conservative foreground strategy that has developed alongside this work is to discard $k$--modes which could be contaminated ({\it avoidance}). The more aggressive alternative is to subtract a sky model and regain access to modes lost by {\it avoidance}. In both cases, which parts of the sky are most critical to either {\it avoid} or {\it subtract} has remained largely uncertain. Here, we will focus primarily on extending the {\it avoidance} strategy by identifying foreground components at greatest risk to ``leak'' from foreground modes to EoR modes and demonstrating a scheme for down--weighting these components.

@@

Interferometer array data known as {\it visibilities}, $V_f(\overline{\mathbf{u}},f)$, represent correlations between time--series of electric fields measured by different antenna pairs with separation vectors $\overline{\mathbf{x}}$ and then Fourier transformed along the time axis to obtain a spectrum along the frequency ($f$) axis. Here, $\overline{\mathbf{u}}\equiv \overline{\mathbf{x}}f/c \equiv \overline{\mathbf{x}}/\lambda$, $c$ is the speed of light, and $\lambda$ is the wavelength of radio waves. $\overline{\mathbf{u}}$ is related to the transverse spatial frequency modes ($\overline{\mathbf{k}}_\perp$) of the brightness distribution of the sky as $\overline{\mathbf{k}}_\perp \equiv 2\pi\overline{\mathbf{u}}/D(z)$, where $D(z)$ is the transverse comoving distance at redshift $z$. If $I(\overline{\mathbf{l}},f)$ is the emission at different frequencies on the sky as a function of direction unit vector ($\overline{\mathbf{l}}$) specified as direction cosines, $V_f(\overline{\mathbf{u}},f)$ represents the Fourier decomposition of $I(\overline{\mathbf{l}},f)$ attenuated by the antenna's angular power--pattern, $A(\overline{\mathbf{l}},f)$, into transverse spatial frequency modes. Taking into account the instrumental bandpass weights $W_f(f)$, this relation can be written as:
\begin{align}\label{eqn:obsvis}
  V_f(\overline{\mathbf{u}},f) &= \iint\limits_\textrm{sky} A(\overline{\mathbf{l}},f)\,I(\overline{\mathbf{l}},f)\,W_f(f)\,e^{-i2\pi \overline{\mathbf{u}}\cdot\overline{\mathbf{l}}}\,\dif^2 \overline{\mathbf{l}}.
\end{align}

We define the {\it delay spectrum}, $V_\eta(\overline{\mathbf{u}},\eta)$, to be the inverse Fourier transform of $V_f(\overline{\mathbf{u}},f)$ along the frequency coordinate:
\begin{align}\label{eqn:delay-transform}
  V_\eta(\overline{\mathbf{u}},\eta) &= \int V_f(\overline{\mathbf{u}},f)\,W'_f(f)\,e^{i2\pi\eta f}\,\dif f,
\end{align}
where, $W'_f(f)$ is a frequency window weighting function which can be chosen to control the quality of the delay spectrum \citep{thy13,ved12} and $\eta$ represents the signal delay between antenna pairs:
\begin{equation}\label{eqn:delay}
  \eta = \frac{\overline{\mathbf{x}}\cdot\overline{\mathbf{l}}}{c} = \frac{\overline{\mathbf{u}}\cdot\overline{\mathbf{l}}}{f}.
\end{equation}
Note that as the visibility is a complex quantity, its Fourier transform is Nyquist sampled over positive and negative delays. Negative and positive delay\footnote{The terms delay and lag are used interchangeably in this paper.} correspond to the two hemispheres of the sky as transected by the antenna spacing vector. In this paper, $\overline{\mathbf{u}}$ and $\overline{\mathbf{x}}$ are assumed to be on a coordinate system aligned with the local east, north (along local meridian) and upward directions at the telescope site. Hence, a perfectly eastward oriented antenna spacing will observe an object in the eastern sky at positive delay, and conversely for an object in the western sky. Similarly, an object in the northern sky will appear at a positive delay at an antenna spacing oriented northward.

Since we are studying a redshifted H{\sc i} spectral line, $f$ is a measure of cosmological distance along the line of sight. Consequently, $\eta$ is a measure of line--of--sight spatial frequency mode given by $k_\parallel\approx 2\pi\eta\,f_{21}H_0\,E(z)/(c(1+z)^2)$, where $f_{21}$ is the rest frame frequency of the 21~cm spin flip transition of H{\sc i}, and $H_0$ and $E(z)\equiv [\Omega_\textrm{M}(1+z)^3+\Omega_\textrm{k}(1+z)^2+\Omega_\Lambda]^{1/2}$ are standard terms in cosmology. We use $H_0=100\,h$~km~s$^{-1}$~Mpc$^{-1}$ and assume a flat $\Lambda$CDM universe with $\Omega_\textrm{M}=0.27$ and $\Omega_\Lambda=0.73$.

Thus, $|V_\eta(\overline{\mathbf{u}},\eta)|^2$ is directly related to the spatial power spectrum of EoR H{\sc i} distribution, $|V(\overline{\mathbf{k}}_\perp,k_\parallel)|^2$ \citep{mor04}. However, foreground objects also emit at radio frequencies. Due to their spatial and spectral (chromatic) structure, the measured $V_\eta(\overline{\mathbf{u}},\eta)$ is contaminated by foreground radio objects \citep{thy13,tro12,mor12,bow09}. These contaminations have to be characterized precisely in order to reduce their impact on EoR H{\sc i} power spectrum detection sensitivity.

In summary, $V_\eta(\overline{\mathbf{u}},\eta)$ is closely related to visibilities, $V_f(\overline{\mathbf{u}},f)$, the basic data blocks measured by each antenna pair, as seen from equation~\ref{eqn:delay-transform}. $V_\eta(\overline{\mathbf{u}},\eta)$ captures all the effects of EoR H{\sc i} signal corruption caused by foregrounds and the instrument. At the same time, it is directly related to the sought power $|V(\overline{\mathbf{k}}_\perp,k_\parallel)|^2$ in $k$--modes containing critical information about spatial scales. For our analysis, we use an equivalent quantity, $V_\eta(\overline{\mathbf{x}},\eta) \equiv V_\eta(\lambda\overline{\mathbf{u}},\eta)$. For purposes of illustration, we often use a simplified quantity, $|V_\eta(|\overline{\mathbf{x}}|,\eta)|$.

Figure~\ref{fig:fourier-space} illustrates the Fourier space in which the delay (and power) spectra of H{\sc i} from the EoR are presented in this paper. $|\overline{\mathbf{x}}|$ and $k_\perp$, denoting spatial scales in the transverse direction, form the $x$--axis. $\eta$ and $k_\parallel$, denoting spatial scales along line--of--sight form the $y$--axis. Foreground emission maps to a wedge--shaped region in Fourier space, hereafter referred to as the {\it foreground wedge}, whose boundaries are determined by antenna spacing and the light travel time across it. These boundaries are called horizon delay limits and are shown by solid lines. The spectral transfer function of the instrument convolves the {\it foreground wedge} and stretches it further (unshaded narrow strips bounded by solid and dashed lines) along $\eta$--axis. This effect has been quantified in \citet{thy13}. The width of this narrow strip is inversely proportional to the operating bandwidth. The region of Fourier space excluding the {\it foreground wedge} and the narrow strips is the so--called {\it EoR window}, shown in light and medium shades. However, the instrument used in our study, namely the MWA, has a passband constructed using coarse channels with a characteristic shape giving rise to periodic gratings thereby resulting in repetitions of the {\it foreground wedge} at multiples of 0.78~$\mu$s. The underlying cause of this effect will be discussed later in this paper.

% Hence, we have identified a region of high significance (shown in medium shade) that forms a subset of the {\it EoR window} bounded by $|\eta|\le 0.78\,\mu$s where contamination from foregrounds is expected to be less and yet the EoR H{\sc i} power spectrum signal is reasonably strong. Hereafter, we refer to this region as the {\it EoR window} in the rest of this paper.

In the context of EoR studies in Fourier space, the H{\sc i} power spectrum from the EoR is expected to decrease rapidly with $|\overline{\mathbf{k}}|$. Hence, the brightest EoR signal will be observed on the shortest baselines and smallest delays. Thus the region of interest for EoR studies is slightly beyond the horizon delay limits (dashed lines) and inside the first grating on either side ($|\eta| \lesssim 0.78\,\mu$s, dot--dashed lines) caused by the instrument's spectral transfer function. We call this region with medium shade as the {\it EoR window} in this paper. The MWA EoR collaboration actively uses this as a test region in Fourier space for diagnostics on data quality.

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/fourier_space_185.0_MHz_30.7_MHz.eps}
\caption{Illustration of Fourier space in which delay (and power) spectra of H{\sc i} signals from the EoR are presented. The $x$--axis is denoted by $|\overline{\mathbf{x}}|$ (antenna spacing) or $k_\perp$ (transverse wavenumber). The $y$--axis denoted by $\eta$ (delay) or $k_\parallel$ (line--of--sight wavenumber). The dark shaded region is referred to as the {\it foreground wedge} where smooth spectrum foregrounds reside. Its boundaries (solid lines), given by light travel time for corresponding antenna spacings, are referred to as horizon delay limits. Regions outside the {\it wedge} and the narrow white unshaded strips (caused by convolution with the instrumental transfer function) are expected to be relatively free of foreground contamination. Due to an undesirable grating pattern (dot--dashed lines) of the instrument used in this study, namely the MWA, we conservatively identify a restricted region of high EoR sensitivity (medium shade) and refer to it as the {\it EoR window} in this paper. \label{fig:fourier-space}}
\end{figure}

@@

The MWA construction was completed in 2012 and, after commissioning, began its EoR observing program in 2013. In its final configuration the MWA is a 128--tile interferometer array capable of observing a 30.72~MHz instantaneous band anywhere in the range 80--300~MHz. Each tile is a phased array of 16 dipoles, each in the shape of a bow--tie. This yields a primary field of view $\gtrsim$~20\arcdeg~ wide and multiple secondary lobes. The array is arranged as a centrally condensed core of $\sim$~300~m --- there are many spacings in the range 5--50~m --- and a radial density that falls off as the inverse of the radius, with the longest baselines at 3~km \citep{bea12}. Here we focus on antenna spacings $|\overline{\mathbf{x}}| \le 200$~m (spatial scales relevant to reionization) and a frequency range 170--200~MHz. The center frequency of 185~MHz corresponds to a redshift, $z\simeq 6.68$.

@@

We obtain the delay spectrum of these calibrated visibilities taking the delay transform of each baseline's spectrum (equation~\ref{eqn:delay-transform}) choosing $W'_f(f)$ to be a {\it Blackman--Harris} window function. The spectrum is multiplied by the additional weight of the flagged channels which occur rarely for interference and regularly every 1.28~MHz where the edges of coarse channels are known to have a small aliasing effect. In delay--space, these weights translate into a convolution by a grated point spread function (PSF) with gratings at multiples of $\eta=0.78\,\mu$s. We deconvolve this PSF using a one dimensional CLEAN algorithm \citep{tay99} as described for the delay axis by \citet{par09,par12}. The CLEAN procedure iteratively finds and subtracts peak values convolved by the Fourier transform of the weights. We limit the selection of peaks to modes inside the horizon delay limit, corresponding to smooth spectrum objects in the visible sky hemisphere.

In subsequent sections, we plot the amplitude of the delay spectrum, $|V(|\overline{\mathbf{x}}|,\eta)|$, sorted by baseline length, $|\overline{\mathbf{x}}|$. Note that visibilities from different baselines have not been averaged together in this analysis.

@@

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/multi_baseline_fhd_delay_spectrum_snapshots.eps}
\caption{Delay spectra of two observations of the EoR field (RA = 0h, Dec = $-30$\arcdeg) at an {\it off--zenith} pointing (top) when tiles have significant secondary lobes in the direction of the Galactic center and a {\it zenith} pointing (bottom) where secondary lobes are minimal and the Galaxy has set. To get the delay spectrum, each baseline is Fourier transformed into delay space ($y$--axis) and the amplitude plotted in order of baseline length ($x$--axis). For reference, the $k$--space equivalent of $\eta$ and $|\overline{\mathbf{x}}|$, namely, $k_\parallel$ and $k_\perp$ are shown at the axes on the right and top respectively. The logarithmic color scale (shown at the right) is common to both panels. White lines mark the boundaries of {\it foreground wedge} determined by light travel time across each baseline. Smooth spectrum objects predominantly reside in this {\it foreground wedge}. The various features in this region depend on antenna spacing orientation, secondary lobe structure of the power--pattern, and the sky at that instant. \label{fig:fhd_data}}
\end{figure}

@@

The instrument model consists primarily of the tile beam power--pattern, $A(\overline{\mathbf{l}},f)$ (see equation~\ref{eqn:obsvis}). It is modeled as a mutually--coupled 4--by--4 dipole array with the overall power--pattern of each individual dipole calculated via finite element electromagnetic simulation \citep{sut14}. To speed up simulation we find that a phased array of isotropic radiators at a height of 0.3~m above an infinite ground plane provides a very good approximation to the full simulation, hence we use the idealized dipoles. We also assume that each individual dipole signal has random delay fluctuations of rms 0.05~ns, a number in line with the known repeatability and stability level of the analog signal chain \citep{bow07b}. The power--patterns are shown in Figure~\ref{fig:power_pattern} for the pointings chosen in our study.

@@

In addition to intrinsic model uncertainty, delay spectra from simulations and data each have fluctuations due to thermal noise in the delay spectrum with rms $\sim$ 1.4$\times 10^7$~Jy~Hz. We estimate the ratio of delay spectra from data and simulations as $\rho = |V^\textrm{D}_\eta(|\overline{\mathbf{x}}|,\eta)|\,/\,|V^\textrm{S}_\eta(|\overline{\mathbf{x}}|,\eta)|$, where superscripts D and S denote data and simulation, respectively. % Using the aforementioned uncertainties added in quadrature, we estimate the approximate expected uncertainty in $\log_{10}\rho$ and denote it by $\Delta\,\log_{10}\rho$.

@@

Figure~\ref{fig:noiseless-dsm-delay-spectrum} shows the amplitude of delay spectrum, $|V_\eta(|\overline{\mathbf{x}}|,\eta)|$, simulated with only the diffuse foreground model, omitting compact sources. The top and bottom panels correspond to {\it off--zenith} and {\it zenith} pointings, respectively.

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/delta_array_multi_baseline_CLEAN_noiseless_visibilities_0.3m_ground_custom_gaussian_FG_model_dsm_all_sky_nside_64_Tsys_95.0K_185.0_MHz_30.7_MHz_bhw2.0.eps}
\caption{Simulated delay spectrum amplitudes $|V_\eta(|\overline{\mathbf{x}}|,\eta)|$ (in units of Jy~Hz) for {\it off--zenith} (top) and {\it zenith} (bottom) pointings for the diffuse foreground model without any thermal noise. White lines mark the boundaries of the {\it foreground wedge} determined by the horizon delay limit and antenna spacing. The axes and color scale are identical to those in Figures~\ref{fig:fhd_data} and \ref{fig:sim_data}. In the {\it zenith} pointing, delay spectrum from diffuse emission has a {\it two--pronged fork}--shaped structure and is present even at wide antenna spacings. \label{fig:noiseless-dsm-delay-spectrum}}
\end{figure}

@@

Figure~\ref{fig:noiseless-dsm-delay-spectrum} also demonstrates diffuse emission from within the primary field of view manifests in the {\it off--zenith} pointing as a branch at the top of the wedge, and at zero delay in the {\it zenith} pointing. The former is seen at positive delays because the primary lobe of the power--pattern is centered eastward of zenith, whereas in the latter it is centered at zenith. As we see from Equation \ref{eqn:obsvis}, each baseline measures a single spatial mode on the sky with an angular size scale inversely proportional to the length of the baseline projected in the direction of the emission. Thus, in the {\it zenith} pointing, the signature of the smooth sky model, the horizontal line at zero delay, fades away on antenna spacings $|\overline{\mathbf{x}}| \gtrsim 125$~m because the sky model is devoid of spatial structures on scales $\lesssim$~0\fdg 75. In the off-zenith pointing, we see power out to much longer baselines which is described below.

@@

A very interesting signature of diffuse emission is revealed at regions near the positive horizon delay limit in the {\it off--zenith} pointing and at both positive and negative horizon delay limits in the {\it zenith} pointing even on wide antenna spacings (125~m~$\lesssim |\overline{\mathbf{x}}|\le$~200~m). This is contrary to expectations because these longer baselines are not expected to be sensitive to diffuse emission (on scales $\gtrsim$~0\fdg 85) of our input foreground model. In the wide--field observations of the MWA, long baselines appear shortened due to projection effects for off--axis emission, especially towards the horizon.  This effect is evident at all LSTs in our simulations and it naturally holds for short baselines, as well. Thus, diffuse emission from far off--axis directions manifests as an edge--heavy {\it two--pronged fork} across all baselines. It decreases in strength with increasing baseline length as expected but is present in all baseline orientations.   %% This contributes to a {\it pitchfork}--shaped feature which will be discussed later in \S\ref{sec:composite} in more detail.

>>>>

Considerable effort has been and continues to be made towards understanding the $k$--space behavior of foreground signatures in the observed power spectrum and formulating robust estimators of the true power spectrum \citep{bow09,liu09,dat10,liu11,mor12,tro12,pob13,thy13,liu14a,liu14b}. A general model of scattering of smooth spectrum foregrounds to higher $k$--modes by the wide-field response of the instrument, into the so-called `wedge', has emerged that provides a high level explanation for the observed foreground power spectrum. The conservative foreground strategy that has developed alongside this work is to discard $k$--modes which could be contaminated ({\it avoidance}). The more aggressive alternative is to subtract a sky model and regain access to modes lost by {\it avoidance}. In both cases, which parts of the sky are most critical to either {\it avoid} or {\it subtract} has remained largely uncertain. Here, we will focus primarily on extending the {\it avoidance} strategy by identifying foreground components at greatest risk to ``leak'' from foreground modes to EoR modes and demonstrating a scheme for down--weighting these components.

@@

Interferometer array data known as {\it visibilities}, $V_{bf}(\vec{b},f)$, represent correlations between time--series of electric fields measured by different antenna pairs with separation vectors $\vec{b}$ and then Fourier transformed along the time axis to obtain a spectrum along the frequency ($f$) axis. If $I(\hat{s},f)$ and $A(\hat{s},f)$ are the sky brightness and antenna's directional power--pattern, respectively, at different frequencies on the sky as a function of direction on the sky denoted by the unit vector ($\hat{s}$), and $W_f(f)$ denotes instrumental bandpass weights, then $V_{bf}(\vec{b},f)$ can be written as (with a slight adaptation from \citet{van34}, \citet{zer38}, and \citet{tho01}):
\begin{align}\label{eqn:obsvis}
  V_{bf}(\vec{b},f) &= \iint\limits_\textrm{sky} A(\hat{s},f)\,I(\hat{s},f)\,W_f(f)\,e^{-i2\pi f\frac{\vec{b}\cdot\hat{s}}{c}}\,\dif\Omega,
\end{align}
where, $c$ is the speed of light, and $\dif\Omega$ is the solid angle element along $\hat{s}$.

We define the {\it delay spectrum}, $V_{b\tau}(\vec{b},\tau)$, to be the inverse Fourier transform of $V_{bf}(\vec{b},f)$ along the frequency coordinate:
\begin{align}\label{eqn:delay-transform}
  V_{b\tau}(\vec{b},\tau) &= \int V_{bf}(\vec{b},f)\,W'_f(f)\,e^{i2\pi f\tau}\,\dif f,
\end{align}
where, $W'_f(f)$ is a spectral weighting function which can be chosen to control the quality of the delay spectrum \citep{thy13,ved12}, and $\tau$ represents the signal delay between antenna pairs:
\begin{equation}\label{eqn:delay}
  \tau = \frac{\vec{b}\cdot\hat{s}}{c}.
\end{equation}

Equation~\ref{eqn:obsvis} can be modified to express the Fourier decomposition of transverse sky brightness distribution as:
\begin{align}\label{eqn:vis}
  V_{uf}(\vec{u},f) &= \iint\limits_\textrm{sky} A(\hat{s},f)\,I(\hat{s},f)\,W_f(f)\,e^{-i2\pi\vec{u}\cdot\hat{s}}\,\dif\Omega,
\end{align}
where, $\vec{u}$ denotes the transverse spatial frequency vector. $\vec{u}$ is directly related to $\vec{b}$ as $\vec{u}\equiv \vec{b}/\lambda$, and to the transverse spatial wavenumber mode, $\vec{k}_\perp$, as $\vec{k}_\perp\equiv 2\pi\vec{u}/D(z)$, where, $D(z)$ is the transverse comoving distance at redshift $z$. Hence, $V_{uf}(\vec{u},f)\equiv V_{bf}(\lambda\vec{u},f)$. It is important to note that $\vec{b}$ is independent of frequency while $\vec{u}$ is not.

Since we are concerned with a redshifted H{\sc i} spectral line from cosmological distances, $f$ is a measure of cosmological distance along the line of sight. $\eta$, which is the Fourier transform dual of $f$, is used to denote the spatial frequency along the line of sight and has units of time. It is directly related to the line--of--sight wavenumber, $k_\parallel\approx 2\pi\eta\,f_{21}H_0\,E(z)/(c(1+z)^2)$, where $f_{21}$ is the rest frame frequency of the 21~cm spin flip transition of H{\sc i}, and $H_0$ and $E(z)\equiv [\Omega_\textrm{M}(1+z)^3+\Omega_\textrm{k}(1+z)^2+\Omega_\Lambda]^{1/2}$ are standard terms in cosmology. Thus,
\begin{align}\label{eqn:los-transform}
  V_{u\eta}(\vec{u},\eta) &= \int V_{uf}(\vec{u},f)\,W'_f(f)\,e^{i2\pi f\eta}\,\dif f
\end{align}
represents the true spatial Fourier representation of the three--dimensional sky brightness distribution. This representation has been discussed in detail in \citet{mor04}. The spatial power spectrum of EoR H{\sc i} distribution, $P(\vec{k}_\perp,k_\parallel)$, and $V_{u\eta}(\vec{u},\eta)$ are related by $P(\vec{k}_\perp,k_\parallel)\propto |V_{u\eta}(\vec{u},\eta)|^2$.

While $V_{b\tau}(\vec{b},\tau)$ and $V_{u\eta}(\vec{u},\eta)$ in equations~\ref{eqn:obsvis} and \ref{eqn:vis} resemble each other very closely, they are not quite identical. The difference basically arises from the fact noted above that $\vec{b}$ is independent of frequency while $\vec{u}$ is not. \citet{par12} and \citet{liu14a} have discussed the mathematical correspondence between the two quantities.

Radio emission from foreground objects, due to their spatial and spectral (chromatic) structure, contaminate the received signal \citep{bow09,mor12,tro12,thy13,liu14a,liu14b}. Since this contamination is expected to be several orders of magnitude stronger than the underlying EoR H{\sc i} signal, it is critical to characterize foregrounds precisely in order to reduce their impact on EoR H{\sc i} power spectrum detection sensitivity. Foregrounds can be described in both the delay spectrum and Fourier frameworks. For our study, which primarily concerns with foreground characterization, we find the former to be simple and yet extremely useful, while maintaining a close correspondence with the latter despite their subtle differences.

In summary, the delay spectrum, $V_{b\tau}(\vec{b},\tau)$, is obtained from {\it visibilities} which are the basic data blocks measured by each antenna pair, using equations~\ref{eqn:obsvis} and \ref{eqn:delay-transform}. $V_{b\tau}(\vec{b},\tau)$ captures all the effects of EoR H{\sc i} signal corruption caused by foregrounds and the instrument. At the same time, it is closely related to the sought power $|V_{u\eta}(\vec{u},\eta)|^2$ in Fourier modes containing critical information about spatial scales. For purposes of illustration later in this paper, we often use the quantity, $|V_{b\tau}(|\vec{b}|,\tau)|$.

Note that since {\it visibility} is a complex quantity, its Fourier transform is Nyquist sampled over positive and negative delays. Negative and positive delay\footnote{The terms delay and lag are used interchangeably in this paper.} correspond to the two hemispheres of the sky as transected by the antenna spacing vector. In this paper, $\vec{b}$ is assumed to be on a coordinate system aligned with the local east, north (along local meridian) and upward directions at the telescope site. Hence, a perfectly eastward oriented antenna spacing will observe an object in the eastern sky at positive delay, and conversely for an object in the western sky. Similarly, an object in the northern sky will appear at a positive delay at an antenna spacing oriented northward.

Figure~\ref{fig:fourier-space} illustrates the Fourier space in which the delay (and power) spectra of H{\sc i} from the EoR are presented in this paper. $|\vec{b}|$ and $k_\perp$, denoting spatial scales in the transverse direction, form the $x$--axis. $\tau\approx\eta\propto k_\parallel$, denoting spatial scales along line--of--sight form the $y$--axis. Foreground emission maps to a wedge--shaped region in Fourier space, hereafter referred to as the {\it foreground wedge}, whose boundaries are determined by antenna spacing and the light travel time across it. These boundaries are called horizon delay limits and are shown by solid lines. The spectral transfer function of the instrument convolves the {\it foreground wedge} and stretches it further (unshaded narrow strips bounded by solid and dashed lines) along $\tau$--axis. This effect has been quantified in \citet{thy13}. The width of this narrow strip is inversely proportional to the operating bandwidth. The region of Fourier space excluding the {\it foreground wedge} and the narrow strips is the so--called {\it EoR window}, shown in light and medium shades. However, the instrument used in our study, namely the MWA, has a passband constructed using coarse channels with a characteristic shape giving rise to periodic gratings thereby resulting in repetitions of the {\it foreground wedge} at multiples of 0.78~$\mu$s. The underlying cause of this effect will be discussed later in this paper.

% Hence, we have identified a region of high significance (shown in medium shade) that forms a subset of the {\it EoR window} bounded by $|\tau|\le 0.78\,\mu$s where contamination from foregrounds is expected to be less and yet the EoR H{\sc i} power spectrum signal is reasonably strong. Hereafter, we refer to this region as the {\it EoR window} in the rest of this paper.

In the context of EoR studies in Fourier space, the H{\sc i} power spectrum from the EoR is expected to decrease rapidly with $|\vec{k}|$. Hence, the brightest EoR signal will be observed on the shortest baselines and smallest delays. Thus the region of interest for EoR studies is slightly beyond the horizon delay limits (dashed lines) and inside the first grating on either side ($|\tau| \lesssim 0.78\,\mu$s, dot--dashed lines) caused by the instrument's spectral transfer function. We call this region with medium shade as the {\it EoR window} in this paper. The MWA EoR collaboration actively uses this as a test region in Fourier space for diagnostics on data quality.

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/fourier_space_185.0_MHz_30.7_MHz.eps}
\caption{Illustration of Fourier space in which delay (and power) spectra of H{\sc i} signals from the EoR are presented. The $x$--axis is denoted by $|\vec{b}|$ (antenna spacing) or $k_\perp$ (transverse wavenumber). The $y$--axis denoted by $\tau$ (delay) or $k_\parallel$ (line--of--sight wavenumber). The dark shaded region is referred to as the {\it foreground wedge} where smooth spectrum foregrounds reside. Its boundaries (solid lines), given by light travel time for corresponding antenna spacings, are referred to as horizon delay limits. Regions outside the {\it wedge} and the narrow white unshaded strips (caused by convolution with the instrumental transfer function) are expected to be relatively free of foreground contamination. Due to an undesirable grating pattern (dot--dashed lines) of the instrument used in this study, namely the MWA, we conservatively identify a restricted region of high EoR sensitivity (medium shade) and refer to it as the {\it EoR window} in this paper. \label{fig:fourier-space}}
\end{figure}

@@

The MWA construction was completed in 2012 and, after commissioning, began its EoR observing program in 2013. In its final configuration the MWA is a 128--tile interferometer array capable of observing a 30.72~MHz instantaneous band anywhere in the range 80--300~MHz. Each tile is a phased array of 16 dipoles, each in the shape of a bow--tie. This yields a primary field of view $\gtrsim$~20\arcdeg~ wide and multiple secondary lobes. The array is arranged as a centrally condensed core of $\sim$~300~m --- there are many spacings in the range 5--50~m --- and a radial density that falls off as the inverse of the radius, with the longest baselines at 3~km \citep{bea12}. Here we focus on antenna spacings $|\vec{b}| \le 200$~m (spatial scales relevant to reionization) and a frequency range 170--200~MHz. The center frequency of 185~MHz corresponds to a redshift, $z\simeq 6.68$.

@@

We obtain the delay spectrum of these calibrated visibilities taking the delay transform of each baseline's spectrum (equation~\ref{eqn:delay-transform}) choosing $W'_f(f)$ to be a {\it Blackman--Harris} window function. The spectrum is multiplied by the additional weight of the flagged channels which occur rarely for interference and regularly every 1.28~MHz where the edges of coarse channels are known to have a small aliasing effect. In delay--space, these weights translate into a convolution by a grated point spread function (PSF) with gratings at multiples of $\tau=0.78\,\mu$s. We deconvolve this PSF using a one dimensional CLEAN algorithm \citep{tay99} as described for the delay axis by \citet{par09,par12}. The CLEAN procedure iteratively finds and subtracts peak values convolved by the Fourier transform of the weights. We limit the selection of peaks to modes inside the horizon delay limit, corresponding to smooth spectrum objects in the visible sky hemisphere.

In subsequent sections, we plot the amplitude of the delay spectrum, $|V_{b\tau}(|\vec{b}|,\tau)|$, sorted by baseline length, $|\vec{b}|$. Note that visibilities from different baselines have not been averaged together in this analysis.

@@

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/multi_baseline_fhd_delay_spectrum_snapshots.eps}
\caption{Delay spectra of two observations of the EoR field (RA = 0h, Dec = $-30$\arcdeg) at an {\it off--zenith} pointing (top) when tiles have significant secondary lobes in the direction of the Galactic center and a {\it zenith} pointing (bottom) where secondary lobes are minimal and the Galaxy has set. To get the delay spectrum, each baseline is Fourier transformed into delay space ($y$--axis) and the amplitude plotted in order of baseline length ($x$--axis). For reference, the $k$--space equivalent of $\tau$ and $|\vec{b}|$, namely, $k_\parallel$ and $k_\perp$ are shown at the axes on the right and top respectively. The logarithmic color scale (shown at the right) is common to both panels. White lines mark the boundaries of {\it foreground wedge} determined by light travel time across each baseline. Smooth spectrum objects predominantly reside in this {\it foreground wedge}. The various features in this region depend on antenna spacing orientation, secondary lobe structure of the power--pattern, and the sky at that instant. \label{fig:fhd_data}}
\end{figure}

@@

The instrument model consists primarily of the tile beam power--pattern, $A(\hat{s},f)$ (see equation~\ref{eqn:obsvis}). It is modeled as a mutually--coupled 4--by--4 dipole array with the overall power--pattern of each individual dipole calculated via finite element electromagnetic simulation \citep{sut14}. To speed up simulation we find that a phased array of isotropic radiators at a height of 0.3~m above an infinite ground plane provides a very good approximation to the full simulation, hence we use the idealized dipoles. We also assume that each individual dipole signal has random delay fluctuations of rms 0.05~ns, a number in line with the known repeatability and stability level of the analog signal chain \citep{bow07b}. The power--patterns are shown in Figure~\ref{fig:power_pattern} for the pointings chosen in our study.

@@

In addition to intrinsic model uncertainty, delay spectra from simulations and data each have fluctuations due to thermal noise in the delay spectrum with rms $\sim$ 1.4$\times 10^7$~Jy~Hz. We estimate the ratio of delay spectra from data and simulations as $\rho = |V^\textrm{D}_\tau(|\vec{b}|,\tau)|\,/\,|V^\textrm{S}_\tau(|\vec{b}|,\tau)|$, where superscripts D and S denote data and simulation, respectively. % Using the aforementioned uncertainties added in quadrature, we estimate the approximate expected uncertainty in $\log_{10}\rho$ and denote it by $\Delta\,\log_{10}\rho$.

@@

Figure~\ref{fig:noiseless-dsm-delay-spectrum} shows the amplitude of delay spectrum, $|V_{b\tau}(|\vec{b}|,\tau)|$, simulated with only the diffuse foreground model, omitting compact sources. The top and bottom panels correspond to {\it off--zenith} and {\it zenith} pointings, respectively.

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/delta_array_multi_baseline_CLEAN_noiseless_visibilities_0.3m_ground_custom_gaussian_FG_model_dsm_all_sky_nside_64_Tsys_95.0K_185.0_MHz_30.7_MHz_bhw2.0.eps}
\caption{Simulated delay spectrum amplitudes $|V_{b\tau}(|\vec{b}|,\tau)|$ (in units of Jy~Hz) for {\it off--zenith} (top) and {\it zenith} (bottom) pointings for the diffuse foreground model without any thermal noise. White lines mark the boundaries of the {\it foreground wedge} determined by the horizon delay limit and antenna spacing. The axes and color scale are identical to those in Figures~\ref{fig:fhd_data} and \ref{fig:sim_data}. In the {\it zenith} pointing, delay spectrum from diffuse emission has a {\it two--pronged fork}--shaped structure and is present even at wide antenna spacings. \label{fig:noiseless-dsm-delay-spectrum}}
\end{figure}

@@

Figure~\ref{fig:noiseless-dsm-delay-spectrum} also demonstrates diffuse emission from within the primary field of view manifests in the {\it off--zenith} pointing as a branch at the top of the wedge, and at zero delay in the {\it zenith} pointing. The former is seen at positive delays because the primary lobe of the power--pattern is centered eastward of zenith, whereas in the latter it is centered at zenith. As we see from Equation \ref{eqn:obsvis}, each baseline measures a single spatial mode on the sky with an angular size scale inversely proportional to the length of the baseline projected in the direction of the emission. Thus, in the {\it zenith} pointing, the signature of the smooth sky model, the horizontal line at zero delay, fades away on antenna spacings $|\vec{b}| \gtrsim 125$~m because the sky model is devoid of spatial structures on scales $\lesssim$~0\fdg 75. In the off-zenith pointing, we see power out to much longer baselines which is described below.

@@

A very interesting signature of diffuse emission is revealed at regions near the positive horizon delay limit in the {\it off--zenith} pointing and at both positive and negative horizon delay limits in the {\it zenith} pointing even on wide antenna spacings (125~m~$\lesssim |\vec{b}|\le$~200~m). This is contrary to expectations because these longer baselines are not expected to be sensitive to diffuse emission (on scales $\gtrsim$~0\fdg 85) of our input foreground model. In the wide--field observations of the MWA, long baselines appear shortened due to projection effects for off--axis emission, especially towards the horizon.  This effect is evident at all LSTs in our simulations and it naturally holds for short baselines, as well. Thus, diffuse emission from far off--axis directions manifests as an edge--heavy {\it two--pronged fork} across all baselines. It decreases in strength with increasing baseline length as expected but is present in all baseline orientations.   %% This contributes to a {\it pitchfork}--shaped feature which will be discussed later in \S\ref{sec:composite} in more detail.
